Bioinformatics analysis of whole-genome bisulfite-sequencing data

### 2015-04-21 Rmarkdown/comethylation.Rmd
Substitute ‘base’ for ‘nucleotide’

Another complexity that these analyses haved brushed over is the fact that the genome is not a 1D line but is rather a complex 3D structure whose shape also varies in time. Therefore, the relevant distance between two loci is not necessarily the number of nucleotides between them but may instead be the Euclidean distance in 3D space. Such complexities cannot be resolved with bisulfite-sequencing data alone but will require measures of the 3D structure of the genome, such as those provided by chromatin conformation capture technologies \citep{Dekker:2013hi}.

>>>>

Another complexity that these analyses haved brushed over is the fact that the genome is not a 1D line but is rather a complex 3D structure whose shape also varies in time. Therefore, the relevant distance between two loci is not necessarily the number of basepair between them but may instead be the Euclidean distance in 3D space. Such complexities cannot be resolved with bisulfite-sequencing data alone but will require measures of the 3D structure of the genome, such as those provided by chromatin conformation capture technologies \citep{Dekker:2013hi}.

### 2015-04-21 Rmarkdown/comethylation_review.Rmd
Fix formatting
Substitute ‘base’ for ‘nucleotide’

In fact, there are well known periodicities in the distribution of nucleotides in genomes. For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ bp period and the latter the $3$ bp period for exonic CpGs.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star.pdf}
\caption{Autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \em{FF} sample.}
\label{fig:FF_chr21_acf_beta_star}
\end{figure}

Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}). Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of nucleotides in the genome than anything to do with their methylation state.

@@

[^cokus_clarification]: A read may contain multiple pairs of CHH loci and it is not clear whether \citeauthor{Cokus:2008fc} consider all pairs of CHH loci or only the first pair of CHH loci in a read. The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci. Either method should give similar results since there will be few 31 nucleotide reads containing multiple pairs of CHH loci.

>>>>

In fact, there are well known periodicities in the distribution of bases in genomes. For example, in eukaryotic genomes there is a $10$ bp period of G+C dinucleotides that acts out of phase with a $10$ bp period of A+T dinucleotides for nucleosome-bound DNA \citep{Segal:2006iz} and a "well-known period-3 oscillation in coding regions" of the genome \citep{Li:1997ts}; the former likely explains the $8$ to $10$ bp period and the latter the $3$ bp period for exonic CpGs.

@@

\begin{figure}[!htb]
\centering
\includegraphics[width=0.8\textwidth]{../figures/FF_chr21_acf_beta_star.pdf}
\caption{Autocorrelation of $\bm{\beta^{\ast}}$ for the chromosome 21 data of the \emph{FF} sample.}
\label{fig:FF_chr21_acf_beta_star}
\end{figure}

Furthermore, if we compute the Fourier transform of this autocorrelation we find a period of approximately 3 bp, identical to that reported by \citet{Lyko:2010dr} for honey bees (compare panel (c) of Figure \ref{fig:Lyko} and Figure \ref{fig:FF_chr21_acf_beta_star_fourier}). Like my earlier criticism of the co-methylation results of \citet{Lister:2009hy}, I believe the apparent $3$ bp period is more likely due to the spacing of bases in the genome than anything to do with their methylation state.

@@

[^cokus_clarification]: A read may contain multiple pairs of CHH loci and it is not clear whether \citeauthor{Cokus:2008fc} consider all pairs of CHH loci or only the first pair of CHH loci in a read. The number of pairs of CHH loci that occur within the same read is not the same as the number of reads containing multiple CHH loci. Either method should give similar results since there will be few $31$ bp reads containing multiple pairs of CHH loci.

### 2015-04-21 Rmarkdown/wgbs_bioinformatics_analysis.Rmd
Minor formatting
Substitute ‘base’ for ‘nucleotide’

The `FastQC` software is a very useful tool for performing this first step ([http://www.bioinformatics.babraham.ac.uk/projects/fastqc/](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)). It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination. `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data. This means that some of its output must be interpreted with caution for bisulfite-sequencing data. For example, `FastQC` will report a warning (resp. error) if the relative frequency of the four nucleotides differ by more than $10\%$ (resp. $20\%$). As noted in the `FastQC` documentation, such a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.

@@

Prior to methylation calling, each read should be filtered to remove low-quality reads and low quality bases. When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from methylation calling[^read_filters]. Strictly speaking each sequenced nucleotide is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all nucleotides within 'surviving' reads. In my own work analysing whole-genome bisulfite-sequencing data, I routinely use the following filters.

@@

The gold-standard is to perform whole-genome DNA sequencing of each sample. This data is then used to form a set of sample-specific methylation loci. However, this approach is also very expensive due to the extra sequencing requirements. A cheaper alternative is to genotype the sample on a genome-wide SNP microarray. This will give very accurate, very cheap genotypes at a large number of loci ($500,000--5,000,000$). However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.

@@

Certain genetic variants, in particular heterozygous $C$>$T$ SNPs, are more difficult to accurately genotype than others. Unfortunately, $C$>$T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants. Fortunately, it is often possible to distinguish $C$>$T$ SNPs from unmethylated cytosines by examining the nucleotide on the opposing strand; if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}). Other base substitutions are more readily detectable, and insertion and deletion events (indels) may also be called.

@@

1. Reference-specific methylation loci, i.e., cytosines in the reference genome that are mutated to non-cytosine nucleotides in the sample's genome.
2. Sample-specific methylation loci, i.e., cytosines in the sample's genome that are non-cytosine nucleotides in the reference genome.
3. Other genetic variants that may be used in additional analyses, such as in identifying allele-specific methylation, or to refine the methylation _type_ or _context_ (see below).

@@

This is done by examining the two nucleotides upstream of the cytosine. It can be done based on the reference sequence, as is done in `Bismark` and `methtuple`, or from the reads themselves. The obvious difficulty with using the reads themselves is if the cytosine occurs at the last or second last position of the read, in which case the context may not be unambiguously determined from the the read alone. Instead, the context may be refined by initially using the reference genome context and then correcting for any sample-specific genetic variants in the two downstream bases.

A further complication occurs when there is a genetic variant in the two nucleotides upstream of the methylation locus. We would like to use the two upstream nucleotides from the sample to infer the sample-specific methylation context, either inferred from each read separately or from a variant calling procedure such as `Bis-SNP`. However, this further complicates the methylation calling and so tools such as `Bismark` derive the context from the reference genome alone.

>>>>

The `FastQC` software is a very useful tool for performing this first step ([http://www.bioinformatics.babraham.ac.uk/projects/fastqc/](http://www.bioinformatics.babraham.ac.uk/projects/fastqc/)). It produces summary graphs of many key measures such as base quality scores, read length distribution and sequence contamination. `FastQC` is a general purpose tool for performing quality control checks of high-throughput sequencing data. This means that some of its output must be interpreted with caution for bisulfite-sequencing data. For example, `FastQC` will report a warning (resp. error) if the relative frequency of the four bases differ by more than $10\%$ (resp. $20\%$). As noted in the `FastQC` documentation, such a warning/error should be ignored for bisulfite-sequencing data, owing to the inherent bias in their sequence composition.

@@

Prior to methylation calling, each read should be filtered to remove low-quality reads and low quality bases. When using a set of filters, at each step a read either 'survives', and is subjected to the proceeding filter, or 'dies', and is excluded from methylation calling[^read_filters]. Strictly speaking each sequenced bases is assigned a weight in the filtering process, however, in practice, filters are normally first applied to reads and then to all bases within 'surviving' reads. In my own work analysing whole-genome bisulfite-sequencing data, I routinely use the following filters.

@@

The gold-standard is to perform whole-genome DNA sequencing of each sample. This data is then used to form a set of sample-specific methylation loci. However, this approach is also very expensive due to the extra sequencing requirements. A cheaper alternative is to genotype the sample on a genome-wide SNP microarray. This will give very accurate, very cheap genotypes at a large number of loci ($500,000$ to $5,000,000$). However, it obviously cannot identify genetic differences that aren't on the array, such as novel sample-specific genetic variants.

@@

Certain genetic variants, in particular heterozygous $C$>$T$ SNPs, are more difficult to accurately genotype than others. Unfortunately, $C$>$T$ SNPs are also quite important because they are the most common SNPs in mammals \citep{Liu:2012ge}, mostly occur at CG dinucleotides and, as a result, are easily mis-called as unmethylated cytosines rather than as genetic variants. Fortunately, it is often possible to distinguish $C$>$T$ SNPs from unmethylated cytosines by examining the base on the opposing strand; if it is a $G$ then the position must be a $C$, if it is a $A$ then it must be a $T$ (see Figure \ref{fig:Bis-SNP}). Other base substitutions are more readily detectable, and insertion and deletion events (indels) may also be called.

@@

1. Reference-specific methylation loci, i.e., cytosines in the reference genome that are mutated to non-cytosine bases in the sample's genome.
2. Sample-specific methylation loci, i.e., cytosines in the sample's genome that are non-cytosine bases in the reference genome.
3. Other genetic variants that may be used in additional analyses, such as in identifying allele-specific methylation, or to refine the methylation _type_ or _context_ (see below).

@@

This is done by examining the two bases upstream of the cytosine. It can be done based on the reference sequence, as is done in `Bismark` and `methtuple`, or from the reads themselves. The obvious difficulty with using the reads themselves is if the cytosine occurs at the last or second last position of the read, in which case the context may not be unambiguously determined from the the read alone. Instead, the context may be refined by initially using the reference genome context and then correcting for any sample-specific genetic variants in the two downstream bases.

A further complication occurs when there is a genetic variant in the two bases upstream of the methylation locus. We would like to use the two upstream bases from the sample to infer the sample-specific methylation context, either inferred from each read separately or from a variant calling procedure such as `Bis-SNP`. However, this further complicates the methylation calling and so tools such as `Bismark` derive the context from the reference genome alone.
